# Reproducible Research: Peer Assessment 1

```{r opts, echo=FALSE, results="hide"}
options(scipen = 1, digits = 4)
```

## Loading and preprocessing the data

First we load the data from the working directory (which is assumed to be the
RepData_PeerAssessment1 folder).

```{r load}
if(!file.exists(paste0(getwd(),"/activity.csv"))) {
    unzip(paste0(getwd(),"/activity.zip"))
}

amd <- read.csv(paste0(getwd(),"/activity.csv"), header = TRUE,
                colClass = c("integer","Date","integer"))
```

## What is mean total number of steps taken per day?

We transform the data to get daily statistics for number of steps (sum, mean,
and median).

```{r stats}
library(plyr)
dayStats <- ddply(amd, c("date"), summarize,
                  mean = mean(steps),
                  median = median(steps),
                  total = sum(steps))
```

Using this data, we create a histogram of the total number of steps taken per day.

```{r hist1}
hist(dayStats$total, col="red", main="Total Steps per Day", xlab="Number of Steps")
```

We also give the mean and median number of steps taken per day.

```{r meanmed1}
print(dayStats[,1:3])
```

## What is the average daily activity pattern?

We calculate the average number of steps taken per time interval and create a
time-series.

```{r timeseries1}
intStats <- ddply(amd, "interval", summarize,
                  mean = mean(steps,na.rm=TRUE))
                   

plot(intStats$interval, intStats$mean, type="l", xlab="Time Interval",
     ylab="Avearge Number of Steps", main="Average Number of Steps per Time Interval")
```

The time interval `r intStats$interval[which.max(intStats$mean)]` contains the
most number of steps on average: `r max(intStats$mean)`.

## Imputing missing values

The variable "steps" contains `r sum(is.na(amd$steps))` missing values. Since
values are missing from whole days, we could replace missing values with the time
interval means calculated previously.

```{r missing}
amdNarm <- amd
for (i in 1:length(amdNarm$steps)) {
   if (is.na(amdNarm$steps[i])) {
       amdNarm$steps[i] <- intStats$mean[intStats$interval == amdNarm$interval[i]]
   }
}

dayStatsNarm <- ddply(amdNarm, "date", summarize,
                      mean = mean(steps),
                      median = median(steps),
                      total = sum(steps))
```

As before, we create a histogram of total number of steps taken per day.

```{r hist2}
hist(dayStatsNarm$total, col="blue", main="Total Steps per Day", xlab="Number of Steps")
```

We also give the mean and median number of steps taken per day.

```{r meanmed2}
print(dayStatsNarm[,1:3])
```

Because we are missing whole days, the statistics for each of the missing days
are identical and the medians are a lot higher. Since the new total number of 
steps for these days is `r dayStatsNarm[1,4]` the only change in the histogram 
is the increase in frequency for the range 10,000-15,000.

## Are there differences in activity patterns between weekdays and weekends?

We start by creating a part of the week (pow) factor variable. Since the calculation
takes a bit of time, we cache the results.

```{r pow, cache=TRUE}
for (i in 1:length(amdNarm$date)) {
    if (weekdays(amdNarm$date[i]) == "Saturday" | weekdays(amdNarm$date[i]) ==
            "Sunday") { amdNarm$pow[i] <- "weekend"
    } else { amdNarm$pow[i] <- "weekday" }
}
amdNarm$pow <- as.factor(amdNarm$pow)
```

We then create a panel plot showing the average number of steps taken per time 
interval for weekdays and weekends.

```{r timeseries2}
intStatsPow <- ddply(amdNarm, c("interval","pow"), summarize,
                     mean = mean(steps))

library(lattice)
xyplot(intStatsPow$mean ~ intStatsPow$interval | intStatsPow$pow, type="l", 
       lwd=1.5, layout = c(1,2), main="Average Number of Steps per Time Interval",
       xlab="Time Interval", ylab="Avearge Number of Steps")

```